---
title: "Reproducible Research Assignment - Economic and Public Health Impact By Extreme Weather Events"
author: "Janos Strasszer"
date: "April 1, 2016"
output: html_document
---

# Economic and Public Health Impact By Extreme Weather Events

## Preface
This analysis was created for Reproducible Research Coursre course. All the data used in the analysis is publicly available and automatically downloaded by the provided scripts.

## Synopsis
Various weather events could cause significant damage in human health and properties also.
This analysis answers the following questions:

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Top 3 events that cause the most injuries:
1. Tornado
2. Flood
3. Excessive heat

Top 3 events that cause the most fatalities:
1. Excessive heat
2. Tornado
3. Flash flood

### Across the United States, which types of events have the greatest economic consequences?

Top 3 events that cause the most property damage:
1. Flood
2. Hurricane/Typhoon
3. Storm surge

Top 3 events that cause the most crop damage:
1. Drought
2. Flood
3. Hurricane


## Data processing

The data used in this publication was downloaded from the NOAA Storm Database.

```{r}
# Downloading data if not exists
if (!file.exists("stormdata.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "stormdata.csv.bz2")
}
stormdata <- read.csv("stormdata.csv.bz2")

# Property and Crop damages should be multiplied by the provided exponenents. (E.g. 2.5k = 2.5 * 1000 = 2500)

# Applying exponents
stormdata[stormdata$PROPDMGEXP == 'H',]$PROPDMG <- stormdata[stormdata$PROPDMGEXP == 'H',]$PROPDMG * 10^2
stormdata[stormdata$PROPDMGEXP == 'K',]$PROPDMG <- stormdata[stormdata$PROPDMGEXP == 'K',]$PROPDMG * 10^3
stormdata[stormdata$PROPDMGEXP == 'M',]$PROPDMG <- stormdata[stormdata$PROPDMGEXP == 'M',]$PROPDMG * 10^6
stormdata[stormdata$PROPDMGEXP == 'B',]$PROPDMG <- stormdata[stormdata$PROPDMGEXP == 'B',]$PROPDMG * 10^9

stormdata[stormdata$CROPDMGEXP == 'H',]$CROPDMG <- stormdata[stormdata$CROPDMGEXP == 'H',]$CROPDMG * 10^2
stormdata[stormdata$CROPDMGEXP == 'K',]$CROPDMG <- stormdata[stormdata$CROPDMGEXP == 'K',]$CROPDMG * 10^3
stormdata[stormdata$CROPDMGEXP == 'M',]$CROPDMG <- stormdata[stormdata$CROPDMGEXP == 'M',]$CROPDMG * 10^6
stormdata[stormdata$CROPDMGEXP == 'B',]$CROPDMG <- stormdata[stormdata$CROPDMGEXP == 'B',]$CROPDMG * 10^9

# Creating aggregated heath damage column (injuries + fatalities)
stormdata$HEALTHDMG <- stormdata$FATALITIES + stormdata$INJURIES

# Adding year column
stormdata$YEAR <- as.integer(format(as.Date(stormdata$BGN_DATE, "%m/%d/%Y"), "%Y"))
```

### Checking avilibity of data

```{r}
options(scipen=5)
hist(stormdata$YEAR, xlab = "Year", ylab = "Frequency of events", main = "Distribution of logged events between 1950 and 2010", breaks=24, col = "lightblue")
```

This plot shows that most of the data was logged between 1995 and 2010. This time period will be used For further analysis.


## Results

### What type of events caused the most fatalities and injuries (1995-2010)
```{r}
stormdata95 <- stormdata[stormdata$YEAR >= 1995,]

healthDamageByEvent <- aggregate(HEALTHDMG ~ EVTYPE, stormdata95, FUN=sum)
healthDamageByEvent <- healthDamageByEvent[rev(order(healthDamageByEvent$HEALTHDMG)),]
healthDamageByEvent$EVTYPE <- factor(healthDamageByEvent$EVTYPE, levels = healthDamageByEvent$EVTYPE)

fatalitiesByEvent <- aggregate(FATALITIES ~ EVTYPE, stormdata95, FUN=sum)
fatalitiesByEvent <- fatalitiesByEvent[rev(order(fatalitiesByEvent$FATALITIES)),]
fatalitiesByEvent$EVTYPE <- factor(fatalitiesByEvent$EVTYPE, levels = fatalitiesByEvent$EVTYPE)

injuriesByEvent <- aggregate(INJURIES ~ EVTYPE, stormdata95, FUN=sum)
injuriesByEvent <- injuriesByEvent[rev(order(injuriesByEvent$INJURIES)),]
injuriesByEvent$EVTYPE <- factor(injuriesByEvent$EVTYPE, levels = injuriesByEvent$EVTYPE)
```

### Top 10 events caused health damage (general damage/injuries/fatalities)
```{r}
# Health damage:
print(head(healthDamageByEvent, 10), row.names = FALSE)
# Fatalities:
print(head(fatalitiesByEvent, 10), row.names = FALSE)
# Injuries:
print(head(injuriesByEvent, 10), row.names = FALSE)
```

### Combined comparison of fatal and non-fatal injuries
```{r}
library(ggplot2)
topNCause <- 10

fatalitiesThreshold = fatalitiesByEvent[topNCause,]$FATALITIES
injuriesThreshold = injuriesByEvent[topNCause,]$INJURIES

injuries <- injuriesByEvent[injuriesByEvent$INJURIES >= injuriesThreshold,]$EVTYPE
fatalities <- fatalitiesByEvent[fatalitiesByEvent$FATALITIES >= fatalitiesThreshold,]$EVTYPE
combined <- unique(unlist(list(injuries, fatalities)))

healthDamagePlot <- merge(x=injuriesByEvent[injuriesByEvent$EVTYPE %in% combined,], y=fatalitiesByEvent[fatalitiesByEvent$EVTYPE %in% combined,], by="EVTYPE")
healthDamagePlot <- healthDamagePlot[rev(order(healthDamagePlot$INJURIES)),]

ggplot(healthDamagePlot, aes(EVTYPE)) + geom_bar(aes(y=FATALITIES+INJURIES), stat="identity", fill = "darkorange") + geom_bar(aes(y=FATALITIES), stat="identity", fill = "darkred") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("Event Type") + ylab("Fatalities + Injuries") + ggtitle("Fatalities and injuries caused by Weather events (1995-2010)")

```

### Which type of events had the greatest economic consequences?
```{r}
propdmgByEvent <- aggregate(PROPDMG ~ EVTYPE, stormdata95, FUN=sum)
propdmgByEvent <- propdmgByEvent[rev(order(propdmgByEvent$PROPDMG)),]
propdmgByEvent$EVTYPE <- factor(propdmgByEvent$EVTYPE, levels = propdmgByEvent$EVTYPE)
cropdmgByEvent <- aggregate(CROPDMG ~ EVTYPE, stormdata95, FUN=sum)
cropdmgByEvent <- cropdmgByEvent[rev(order(cropdmgByEvent$CROPDMG)),]
cropdmgByEvent$EVTYPE <- factor(cropdmgByEvent$EVTYPE, levels = cropdmgByEvent$EVTYPE)
```

### Top 10 events caused propery and crop damage (1995-2010)
```{r}
# Property damage:
print(head(propdmgByEvent, 10), row.names = FALSE)
# Crop damage:
print(head(cropdmgByEvent, 10), row.names = FALSE)

#Plot results
p1 <- ggplot(propdmgByEvent[1:10,], aes(EVTYPE, PROPDMG/1000000)) + geom_bar(stat="identity", fill = "darkorange") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("Event Type") + ylab("Property damage [million $]") + ggtitle("Property damage by Weather events [million $] (1995-2010)")

p2 <- ggplot(cropdmgByEvent[1:10,], aes(EVTYPE, CROPDMG/1000000)) + geom_bar(stat="identity", fill = "darkgreen") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("Event Type") + ylab("Crop damage [million $]") + ggtitle("Crop damage by Weather events [million $] (1995-2010)")

multiplot()
library(gridExtra)
grid.arrange(p1, p2, ncol=2)
```

